---
title: "Reproducible Research- Project 1"
author: "Ryan Glenn P. Deguma"
output: html_document
---

```{r setup, include=FALSE}
#This is just precuationary measures that All needed Libraries are installed and loaded.
knitr::opts_chunk$set(echo = TRUE)
if (!require(toOrdinal)) install.packages('toOrdinal',repos = "http://cran.us.r-project.org")
library(toOrdinal)

if (!require(ggplot2)) install.packages('ggplot2',repos = "http://cran.us.r-project.org")
library(ggplot2)
```

## Introduction

This a Markdown Document that describes the procedure conducted to answer the questions given.

### Loading and preprocessing the data
The following code check if the Activity Data is present on the current working directory. If not, it will download the file from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). 

```{r read_data}
activityDataFile <- paste0(getwd(),"/activity.csv")

if(!(file.exists(activityDataFile))){
  fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  tempZipFile <- paste0(getwd(), "/temp.zip")
    
  if (Sys.info()['sysname'] == "Windows"){
    download.file(fileUrl, tempZipFile)
  } else {
    download.file(fileUrl, tempZipFile, method = "curl")
  }
    
  unzip(tempZipFile)
  
  file.remove(tempZipFile)
}

activityData <- read.csv(activityDataFile)
activityData$date <- as.POSIXct(activityData$date, "%Y-%m-%d", tz="")

summary(activityData)

```

### What is mean total number of steps taken per day?
```{r calc_daily_steps}
daily_steps<- aggregate(activityData$steps, by = list(activityData$date), FUN = sum, na.rm = TRUE)
names(daily_steps) <- c("date", "steps")

steps_mean <- format(mean(daily_steps$steps), digits = 2, nsmall = 2, big.mark = ',')
steps_median <- format(median(daily_steps$steps), digits = 2, nsmall = 2, big.mark = ',')

hist(daily_steps$steps, main = "Total Number of Steps Taken Daily", xlab = "Daily Steps", col = "Red")
```

The **mean** and **median** total number of steps taken per day are **`r steps_mean`** and **`r steps_median`** respectively.

### What is the average daily activity pattern?
```{r calc_avg_steps}
average_daily_steps <- aggregate(activityData$steps, by=list(activityData$interval), FUN=mean, na.rm=TRUE)
names(average_daily_steps) <- c("interval", "mean")

steps_max <- average_daily_steps[which.max(average_daily_steps$mean), ]$interval

plot(average_daily_steps$interval, average_daily_steps$mean, type = "l", xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
```

The **`r toOrdinal(steps_max)`** of the 5-minute interval has the maximum average number of steps.

### Imputing missing values
```{r count_na}
noNA <- format(sum(is.na(activityData$steps)), digits = 2, nsmall = 2, big.mark = ',')
```

There are **`r noNA`** missing values in the dataset.   

Replacing the **NA** values with **Mean of the 5-min Interval** computed above and stored in **average_daily_steps**.
```{r Impute_NA}
activityData_Imputed <- transform(activityData, steps = ifelse(is.na(activityData$steps), average_daily_steps$mean[match(activityData$interval,average_daily_steps$interval)], activityData$steps))
```

Now, generate Histogram for the total number of steps taken each day and Calculate its Mean and Median.
```{r Generate_Hist_Impute}
totalDailySteps_Imputed <- aggregate(activityData_Imputed$steps, by = list(activityData_Imputed$date), FUN = sum)
names(totalDailySteps_Imputed) <- c("date", "steps")

stepsImpute_mean <- format(mean(totalDailySteps_Imputed$steps), digits = 2, nsmall = 2, big.mark = ',')
stepsImpute_median <- format(median(totalDailySteps_Imputed$steps), digits = 2, nsmall = 2, big.mark = ',')

hist(totalDailySteps_Imputed$steps, xlab="Daily Steps", main="Total Number of Steps Taken Daily (Imputed Data)", col = "Red")
```

The Mean and Median of the Activity Data with imputed missing values are **`r stepsImpute_mean`** and **`r stepsImpute_median`**. These values are different from the values with missing data.    
In effect, filling out the missing values smooths out the Histogram and move the mean from **`r steps_mean`** to **`r stepsImpute_mean`**.

### Are there differences in activity patterns between weekdays and weekends?
Identifying the Weekday and Weekend, then Plot the Graph
```{r Compare_Wekday}
 activityData_Imputed$DayType <- ifelse(weekdays(activityData_Imputed$date) == "Saturday" | weekdays(activityData_Imputed$date) == "Sunday", "Weekend", "Weekday") 

Mean_ActivityData_Imputed <- aggregate(activityData_Imputed$steps, by = list(activityData_Imputed$interval, activityData_Imputed$DayType), FUN = mean)
names(Mean_ActivityData_Imputed) <- c("Interval", "DayType", "MeanSteps")

plot<- ggplot(Mean_ActivityData_Imputed, aes(x = Interval , y = MeanSteps, color = DayType)) +
       geom_line() +
       labs(title = "Average daily steps by type of day", x = "Interval", y = "Average number of steps") +
       facet_wrap(~DayType, ncol = 1, nrow=2)
print(plot)
```